---
title: "Glycan Biosynthesis Network"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Glycan Biosynthesis Network}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6
)
```

In [Get Started with glyenzy](https://glycoverse.github.io/glyenzy/articles/glyenzy.html),
we briefly discussed using `trace_biosynthesis()` to build glycan biosynthesis networks.
Here, we take a closer look at the process and introduce the concept of a
"virtual enzyme."

```{r setup}
library(glyenzy)
library(glydraw)
```

# `trace_biosynthesis()`

Let's start with the O-GalNAc glycan we met earlier:

```{r out.width = "300px", fig.width = 4, fig.height = 3}
glycan <- "Gal(b1-4)GlcNAc(b1-6)[Gal(b1-3)]GalNAc(a1-"
draw_cartoon(glycan)
```

To build its biosynthesis network, call `trace_biosynthesis()`:

```{r}
path <- trace_biosynthesis(glycan)
path
```

`path` is an `igraph` object representing the biosynthesis network.
It contains the following attributes:

- `name`: node attribute, character, IUPAC-condensed sequences of glycan structures
- `enzyme`: edge attribute, character, gene symbol of the enzymes involved in the enzymatic step
- `step`: edge attribute, integer, the rank of the enzymatic step

An enzymatic step can often be catalyzed by more than one enzyme.
This results in multiple edges between two nodes, with each edge representing an enzyme.
Therefore, the resulting network is a multi-edge network.

You can visualize the network with `plot()`.
For a simpler visualization, the multi-edge network is reduced to a
single-edge network when it is plotted.

```{r}
plot(path)
```

`trace_biosynthesis()` also accepts a vector of glycan structures.
When you provide multiple glycans, it infers a union biosynthesis network
that reaches each structure.

```{r}
glycans <- c(
  "Neu5Ac(a2-3)Gal(b1-3)GalNAc(a1-",
  "Neu5Ac(a2-3)Gal(b1-3)[Neu5Ac(a2-6)]GalNAc(a1-",
  "Gal(b1-3)[Gal(b1-4)GlcNAc(b1-6)]GalNAc(a1-"
)
path <- trace_biosynthesis(glycans)
plot(path)
```

Note that "Neu5Ac(a2-3)Gal(b1-3)GalNAc(a1-" is both a target glycan and
an intermediate glycan.

Now let's look at the arguments.
`enzymes` accepts a character vector of enzyme gene symbols.
By default, `enzymes = NULL`, so all enzymes are used.
You can use this argument to restrict your searching space.

```{r}
enzymes <- c("C1GALT1", "ST3GAL1", "ST6GALNAC1", "GCNT1", "B4GALT1")
path <- trace_biosynthesis(glycans, enzymes)
plot(path)
```

The other two arguments, `max_steps` and `filter`, may be less commonly used.
They let you place additional restrictions on the network inference algorithm.
See the function documentation for details.

We'll return to `max_virtual_steps` later.

# `trace_biosynthesis_virtual()`

`trace_biosynthesis()` is based on existing knowledge of glycan biosynthesis.
However, many glycan structures can be detected even when their biosynthesis
routes remain unknown.
One typical example is the O-GalNAc core 5.

```{r out.width = "300px", fig.width = 4, fig.height = 3}
glycan <- "GalNAc(a1-3)GalNAc(a1-"
draw_cartoon(glycan)
```

The exact enzyme responsible for adding the second GalNAc is still uncertain.
As a result, trying to infer the biosynthesis network for this simple glycan
raises an error.

```{r}
try(trace_biosynthesis(glycan))
```

Sometimes, though, the concrete enzymes matter less than the possible
biosynthesis routes. In that case, `trace_biosynthesis_virtual()` can help.
This function works in an enzyme-agnostic way, building a biosynthesis network
with "virtual enzymes."

```{r fig.width = 3, fig.height = 3}
path <- trace_biosynthesis_virtual(glycan)
plot(path, width = 3, height = 3)
```

Here, "a3GalNAcT" is not a concrete enzyme, but a "virtual enzyme" assigned
to this enzymatic step. In this way, `trace_biosynthesis_virtual()` can infer
a biosynthesis network for any glycan structure.
Let's try it on the three O-GalNAc glycans we saw earlier.

```{r}
path <- trace_biosynthesis_virtual(glycans)
plot(path)
```

Compared with the result from `trace_biosynthesis()`, some new edges appear.
These represent possible, but non-canonical, enzymatic steps.

# Hybrid methods

These two methods are not mutually exclusive.
Both functions provide ways to incorporate the other approach.

For `trace_biosynthesis()`, there is a `max_virtual_steps` argument.
When a residue cannot be assigned to a known enzyme,
`trace_biosynthesis()` can assign it to a virtual enzyme
using the same technique as `trace_biosynthesis_virtual()`.

For example, canonical FUT8 reactions require a b2 GlcNAc added by MGAT1.
This prevents high-mannose N-glycans being core-fucosylated.
Therefore, calling `trace_biosynthesis()` on these glycans raises an error.

```{r}
glycan <- "Man(a1-2)Man(a1-2)Man(a1-3)[Man(a1-2)Man(a1-3)[Man(a1-2)Man(a1-6)]Man(a1-6)]Man(b1-4)GlcNAc(b1-4)[Fuc(a1-6)]GlcNAc(b1-"
try(trace_biosynthesis(glycan))
```

`max_virtual_steps` defines the maximum number of target-directed virtual enzyme
steps allowed when no fully enzymatic path exists.
Setting it to `1L` makes it possible to trace the biosynthesis network for this glycan.

```{r}
path <- trace_biosynthesis(glycan, max_virtual_steps = 1L)
plot(path)
```

Although core fucosylation may occur on any intermediate glycan structure,
the algorithm defers it as much as possible to avoid over-inference.

`trace_biosynthesis_virtual()` also has an `annotate_enzymes` argument.
Setting it to `TRUE` adds a `concrete_enzyme` edge attribute to the resulting network.
Let's try it with the first O-GalNAc glycan:

```{r}
glycan <- "Gal(b1-4)GlcNAc(b1-6)[Gal(b1-3)]GalNAc(a1-"
path <- trace_biosynthesis_virtual(glycan, annotate_enzymes = TRUE)
igraph::edge_attr(path, "concrete_enzymes")
```
